---
title: "Lesson 1"
output: html_document
---

## Lesson 1

File 'sf_crime-data.csv' was downloaded and placed in the same folder as this markdown

```{r crime-wrong}
#read.csv("sf_crime-data-2017.csv") -> not in the correct working directory
```

We need to define the directory properly. This can be done by explicitly defining the file path (absolute) as such:

```{r crime-absolute}
head(read.csv("C:/Users/Carissa Ying/Documents/Year 3 Sem 1/DBA3702/Lesson 1/sf_crime-data-2017.csv"))
```

**However**, this is not recommended since it affects code sharing. We can use relative paths instead.
```{r crime-relative}
# other users can change this part to suit their own systems
setwd("C:/Users/Carissa Ying/Documents/Year 3 Sem 1/DBA3702/Lesson 1")
head(read.csv("sf_crime-data-2017.csv"))
```

```{r crime-relative-2}
# . refers to the current working directory
head(read.csv("./sf_crime-data-2017.csv"))
# .. navigates to the parent directory
```

It is also possible to set the root working directory under Tools > Global Options > General

## Packages

You can install packages by using the 'install.packages("dplyr")' command. You can then load the package using 'library(dplyr)'