---
title: "Lecture 5"
output: html_document
---

## Conditional Statements

```{r}
x = 0

if (x > 0) {
  x = 10
} else if (x < -2) {
  x = 20
} else {
  x = 40
}

x # -2<x<0, returns 40
```

```{r}
ifelse(x<30, "Low", ifelse(x>50, "High", "Medium"))
```

***

## For Loops

```{r}
sum = 0
a = c()
for (i in seq(0, 100, 5)) {
  sum = sum + i
  a = append(a, sum)
}

a
```

``` {r}

properties_data = read.csv("./Data/properties.csv")

nrows = nrow(properties_data)

for (i in 1:nrows) {
  # get the ith property
  property = properties_data[i, ]
  
  # if developer is missing
  if (is.na(property$Developer)) {
    # get other properties with the same name
    idx = properties_data$Property.Name == property$Property.Name &
      !is.na(properties_data$Developer)
    # if they have a developer
    if (sum(idx) > 0) {
      # use data to fill in missing
      property$Developer = properties_data[idx, "Developer"][[1]]
    }
  }
}
```

***

## While Loops

```{r}
fibs = c(1, 1)
i = 1; j = 2

while (fibs[i] + fibs[j] < 1000) { # good for uncertain ranges
  fibs[j+1] = fibs[i] + fibs[j]
  i = i + 1
  j = j + 1
}

fibs
```

Using break to escape the while loop:

```{r}
fibs = c(1, 1)
i = 1; j = 2

while (T) { # good for uncertain ranges
  fibs[j+1] = fibs[i] + fibs[j]
  i = i + 1
  j = j + 1
  if (fibs[i] + fibs[j] >= 1000) { # be careful; can lead to infinite looping
    break
  }
}

fibs
```

***

## Functions

Define a function using:

```{}
function_name = function(inputs) { # can have zero inputs
  actions
  return(output) # optional, returns the last line by default
}
```

```{r}
# will give the tightest fibonacci series from start to stop inclusive
generate_fib = function(stop, start=1) {
  fib = c(1, 1)
  
  i = 1; j = 2;
  while (fib[i] + fib[j] <= stop) {
    fib[j+1] = fib[i] + fib[j]
    i = i + 1
    j = j + 1
  }
  
  fib = fib[fib >= start & fib <= stop]
  return(fib)
}

generate_fib(1000, start=10)
```

***

## Apply Functions

- allows application of a function to a margin of a matrix or data frame without the use of explicit looping
- iterates across rows or columns more efficiently
- apply(), sapply(), mapply(), lapply(), vapply(), rapply(), tapply()

### Apply

```{r}
ID = c(3000, 1234, 2000)
start = c(5, 50, 0) 
end = c(100, 100, 0)

m = cbind(ID, start, end)

# MARGIN = 1 performs operations across rows
# MARGIN = 2 performs operations along columns
# FUNC must be able to accept a variable as its input
apply(m, 2, sum)
```

Can apply user-defined functions too:

```{r}
CountOdd = function(x) {
  return(sum(x%%2))
}

apply(m, 2, CountOdd)

CountOdderEven = function(x, flag) {
  if(flag) {
    return(sum(x%%2)) # flag true, count odd
  } else {
    return(length(x) - sum(x%%2)) # flag false, count even
  }
}

apply(m, 2, CountOdderEven, F) # need to supply the flag variable
```

Other apply functions work similarly but on more specific data structures:

### 2. Lapply

- input is a list of vectors
- returns a list of the same length as the input

```{r}
x = list(A=1:4, B=seq(0.1, 1, by=0.1))

lapply(x, mean)
```

### 3. Sapply

- A wrapper function of lapply
- returns a vector instead of a list

```{r}
x = list(A=1:4, B=seq(0.1, 1, by=0.1))

sapply(x, mean)
```

### 4. Rapply

- recursive apply
- good for lists of lists

```{r}
x = list(A=2, B=list(-1, 3), c=list(-2, list(-5, 6)))

rapply(x, function(x){x^2})
```

### 5. Mapply

- some functions are multivariate
- takes multiple vectors as inputs
- applies functions to combinations of elements from each vector

```{r}
mapply(rep, 1:5, c(4,4,4,4,4))

## 1 3 1
## 2 4 2
A = matrix(c(1,2,3,4), ncol=2) 
## 1 2 3
## 4 5 6
B = matrix(c(1,2,3,4,5,6), ncol=2, byrow=T)
A
B
mapply(rep, A, B)
```

### 6. Tapply

- applies function to each group of an array, grouped by certain factors

```{r}
x = 1:10
y = factor(c("A", "A", "A", "B", "B", "B", "B", "C", "C", "C"))

x
y
tapply(x, y, sum)
```

*** 

## Pivot Table

```{r}
# total murders per region
murders = read.csv("./Data/murders.csv")
tapply(murders$total, murders$region, sum)

# divides population into intervals, gets mean
bins = cut(murders$population, breaks=c(0, 1e+06, 1e+07, 1e+08))
tapply(murders$population, bins, mean)
```

### Split

- splits data frame into a list of data frames by a factor array

```{r}
# ratio of murder cases to population by region
## total by region
total = tapply(murders$total, murders$region, sum)
## population by region
pop = tapply(murders$population, murders$region, sum)
total/pop

# DONE USING SPLIT
region = split(murders, murders$region)
sapply(region, function(x){sum(x$total)/sum(x$population)})
```

### Group by Multiple Factors

Use a list for the aggregating factor:

```{r}
head(mtcars) # built in dataset

# Ave. h-p by a-m and gear
# am becomes the rows, gear becomes the columns
tapply(mtcars$mpg, INDEX=list(mtcars$am, mtcars$gear), mean)
```