---
title: "Tutorial 9"
output: html_document
---

Notes:
- ggmap has geoplot function that can give the longitude and latitude from the city and street name

```{r}
library("dplyr")
library("ggmap")
library("ggplot2")
library("tidyr")
library("lubridate")

data = read.csv("./Data/Police Use of Force.csv")
head(data)
```

1. Convert incident date and time into date time format

```{r}
# detect 4 digit year and 2 digit year
incident_dates = parse_date_time(data$INCIDENT_DATE, c("%-m/%-d/%Y", "%-m/%-d/%y"))
# detect time
incident_times = parse_date_time(data$INCIDENT_TIME, c("%-I:%M:%S %p","%I:%M:%S %p")) 

# combine
incident_dates = incident_dates %>% as_datetime()
incident_times = incident_times %>% as_datetime()
datestrings = incident_dates %>% strftime(format="%d/%m/%Y")

data$NEW_DATETIME = paste(datestrings,incident_times) %>% strptime(format="%d/%m/%Y %H:%M:%S %p", tz="GMT") %>% as.POSIXct()
```

2. Split force-effective column

```{r}
num_cols = data$FORCE_EFFECTIVE %>% 
  sapply(strsplit, ", ") %>%
  sapply(length) %>%
  max()
new_cols = paste("FORCE_EFFECTIVE", seq(1, num_cols), sep="_")
data = separate(data, FORCE_EFFECTIVE, into=new_cols, sep=", ")
```

3. Fill up NA in longitude and latitude using address

```{r}
ggmap::register_google(key='AIzaSyCnGFj3-tmhyhkx2Suxw3HNa6P0c0fjHc0')

missing = is.na(data$LOCATION_LATITUDE)

loc = paste(data$LOCATION_FULL_STREET_ADDRESS_OR_INTERSECTION, data$LOCATION_CITY, ", ")
loc = loc[missing]
loc = geocode(loc, output="latlon")
data[missing, c("LOCATION_LATITUDE", "LOCATION_LONGITUDE")] = loc[c("lat", "lon")]
```

4. Convert number of EC cycles to consistent numbers

```{r}
data$NUMBER_EC_CYCLES = gsub("NULL", NA, data$NUMBER_EC_CYCLES, fixed=T)
data$NUMBER_EC_CYCLES = data$NUMBER_EC_CYCLES %>%
  sapply(strsplit, ", ") %>%
  sapply(as.integer) %>%
  sapply(sum)

head(data)
```

5. Data Visulisation

Racial Discrimination:
- For the offense, compare the severity of the force used between different races
- May also want to compare the officer race

```{r}

```

When experience increases, is officer more likely to shoot:

```{r}
data$OFFICER_YEARS_ON_FORCE = as.integer(data$OFFICER_YEARS_ON_FORCE)

obs = data %>% mutate_at(vars(NUMBER_EC_CYCLES), ~replace(., is.na(.), 0)) %>%
  group_by(OFFICER_YEARS_ON_FORCE) %>% 
  summarise(MEAN_EC_CYCLES = mean(NUMBER_EC_CYCLES > 0))

ggplot(obs, aes(x=OFFICER_YEARS_ON_FORCE, y=MEAN_EC_CYCLES)) + 
  geom_point()
```

Use ggmap/ leaflet to display the distribution of events:

```{r}

```