---
title: "Lecture 7"
output: html_document
---

# Data Visualisation

## Types of Data Visualisation

1. Idea illustration
2. Idea generation
3. Everyday data visualisation
4. Visual discovery
- Trendspotting
- Sense-making
- Deep analysis
  
  
## Principles of Data Visualisation

### 1. Simplify
- captures only the essence of the data

### 2. Compare
- compare visualisations side-by-side
- allow audience to spot trends, patterns and differences by sight

### 3. Attend
- highlight details that audience needs to attend to

### 4. Explore
- Allow audience to explore data and discover things by sight

### 5. View Diversely
- look at the same data from different perspectives at the same time
- learn how the data fits together

### 6. Ask Why
- allow the audience to dig into the data to find out why things are happening

### 7. Be Skeptical
- don't be contented with the first answer we get
- always explore further

### 8. Respond
- share the data to allow others to make use of it and to check it
  

## ggplot2

### 1. Layer
- allow combine graphs in different layers to make one complex graph

### 2. Scale
- maps values in the data space to values in the aesthetic space (e.g. color, size, shape)

### 3. Coordinate system
- how points are positioned in space

### 4. Faceting
- split the data into subsets of the entire data set

```{r, message=FALSE, warning=FALSE}
library("ggplot2")
countries = read.csv("./Data/Countries.csv")
head(countries)
```

### Plotting

Comparing the number of countries among different continents:

```{r}
library("countrycode")
countries$Continent <- countrycode(sourcevar = countries[, "Country"],
                            origin = "country.name",
                            destination = "continent")

ggplot(data=countries,aes(x=Continent)) + geom_bar()
```

**aes** means aesthetics and determines the look of the graph. Here, we are specifying that the Continent column is mapped to the x-axis of the layer. **geom_bar** function means that we are expecting the plot to be a bar chart.

This is only using a few of the possible components of a layer. Each geom function has its own statistical transformation (in this case, count) and position by default. 

We can change this transformation using **stat**.We can also colour the bars using **fill** and specify bar dimensions using **width**.

```{r}
library("dplyr")
df = countries %>% group_by(Continent) %>% summarize(Count=n())

ggplot(data=df,aes(x=Continent, y=Count)) + geom_bar(stat="identity", fill="lightblue", width=0.5)
```

### Overlay Layers

For a line plot, we use **geom_line**. Note that geom_line connects lines in the same group. Hence, we need to specify **group=1**.

```{r}
ggplot(data=df, aes(x=Continent, y=Count)) + geom_line(group=1)
```

We can use **geom_point** for a scatter plot and use **size**, **shape**, and fill to specify the look of our points.

```{r}
ggplot(data=df, aes(x=Continent, y=Count)) + geom_point(size=5, shape=23, fill="blue")
```

To overlay the two, we can just add the two geom functions:

```{r}
ggplot(data=df, aes(x=Continent, y=Count)) + geom_line(group=1) + 
  geom_point(size=5, shape=23, fill="blue")
```

This works in this case because the data here shares the same data source. However, ggplot2 also offers flexibility for plots of different data like so:

```{r}
df = countries %>% group_by(Continent) %>% summarize(Average_GDP_Per_Capita_1000=mean(GDP_Per_Capita)/1000)

ggplot() + geom_bar(data=countries, aes(x=Continent), fill="lightblue", width=0.5) +
  geom_line(data=df, aes(x=Continent, y=Average_GDP_Per_Capita_1000), group=1)
```

Note that the sequence of geom functions determines the order in which the layers are added. Since the geom_line is added last, it is layered on the top.


### Scaling

GDP per Capita vs Life Expectancy:

```{r}
ggplot(data=countries, aes(x=GDP_Per_Capita, y=Life_Expectancy, 
                           size=Population, color=Continent)) + geom_point()
```

This relationship is non-linear (likely logarithmic). We can rescale the GDP per Capita using **scale_x_log10**:

```{r}
ggplot(data=countries, aes(x=GDP_Per_Capita, y=Life_Expectancy, size=Population, 
                           color=Continent)) + geom_point() + scale_x_log10()
```

We can also scale other factors (e.g. point size). In this case, we resize the population to a logarithmic scale using **scale_size_continuous**:

```{r}
ggplot(data=countries, aes(x=GDP_Per_Capita, y=Life_Expectancy, 
                           size=Population, color=Continent)) + 
  geom_point() + scale_x_log10() + scale_size_continuous(trans="log10")
```

We can also set the colours of the points manually using **scale_color_manual**:

```{r}
ggplot(data=countries, aes(x=GDP_Per_Capita, y=Life_Expectancy, 
                           size=Population, color=Continent)) + 
  geom_point() + scale_x_log10() + scale_color_manual(values=c("red", "darkgreen", "purple", "orange", "blue"))
```


### Coordinate System

By default, ggplot2 uses the Cartesian coordinate system. We can switch to a Polar system using **coord_polar**:

```{r}
ggplot(data=countries,aes(x=Continent)) + geom_bar(fill="blue") + coord_polar()
```

### Graph Components

1. Plot title (Text)
2. Axis Title (Text)
3. Tick Label (Text)
4. Background Panel (Rectangle)
5. Grid Lines (Line)
6. Others (found under ?theme)

Each of the 3 types can be manipulated using the **element_text**, **element_rect** and **element_line** functions respectively:

```{r}
## change bg to white
ggplot(data=countries,aes(x=Continent)) + geom_bar(fill="blue") + 
  theme(panel.background=element_rect(fill="white"))

## change orientation of tick labels
ggplot(data=countries,aes(x=Continent)) + geom_bar(fill="blue") + 
  theme(axis.text.x=element_text(size=15, angle=90))
```

We can also adopt a theme:

```{r}
ggplot(data=countries,aes(x=Continent)) + geom_bar(fill="blue") + theme_linedraw()
```

### Faceting

```{r}
library("gapminder")
head(gapminder)
```

gapminder contains the data for multiple years. If we plot it as such, the data becomes very messy. Instead, we can facet the data by year using **facet_wrap** to produce multiple side-by-side plots:

```{r}
ggplot(data=gapminder, aes(x=gdpPercap, y=lifeExp, size=pop, color=continent)) +
  geom_point() + scale_x_log10() + theme_bw() + facet_wrap(~year)
```

### Positioning
- identity (default for most geoms)
- jitter (default for geom_jitter)
- dodge (default of geom_boxplot)
- stack (default of geom_bar, geom_histogram, geom_area)
- fill (useful for geom_bar, geom_histogram, geom_area)

```{r}
custdata = read.csv("./Data/Customers.csv")

ggplot(custdata) + geom_bar(aes(x=marital.status, fill=is.member))
```

```{r}
ggplot(custdata) + geom_bar(aes(x=marital.status, fill=is.member), position="dodge")

```

```{r}
ggplot(custdata) + geom_bar(aes(x=marital.status, fill=is.member), position="fill")
```

## gganimate

Animate the changes in plot by year using **transition_time**

```{r}
library("gganimate")
library("png")
library("gifski")

p = ggplot(data=gapminder, aes(x=gdpPercap, y=lifeExp, size=pop, color=continent)) +
  geom_point() + scale_x_log10() + theme_bw()

p + transition_time(year) + labs(title="Year: {frame_time}")
```

