---
title: "SA1_Carissa Ying Geok Teng"
author: "SA1 Carissa Ying Geok Teng A0205190R"
date: "11/2/2021"
output: html_document
---

```{r}
library("dplyr")
library("lubridate")
library("ggplot2")
library("tidyr")
library("ggmap")
library("leaflet")
```


```{r}
journeys = read.csv("./Data/journeys.csv")
stations = read.csv("./Data/stations.csv")

head(journeys)
head(stations)
```

## Q1

```{r}
journeys = journeys %>% mutate(Start.Time = paste0(Start.Year, "-", 
                                                   Start.Month, "-",
                                                   Start.Date, " ",
                                                   Start.Hour, ":",
                                                   Start.Minute, ":"))

journeys$Start.Time = journeys$Start.Time %>% 
  parse_date_time(c("$Y-%m-%d %H:%M", "%y-%m-%d %H:%M"))
```

## Q2 

```{r}
# Get the end time
journeys = journeys %>% mutate(End.Time = paste0(End.Year, "-", 
                                                   End.Month, "-",
                                                   End.Date, " ",
                                                   End.Hour, ":",
                                                   End.Minute, ":"))

journeys$End.Time = journeys$End.Time %>% 
  parse_date_time(c("$Y-%m-%d %H:%M", "%y-%m-%d %H:%M"))

## Idea: if start time is 14min 59s and time taken is 61s, the end time is 16min 00s. It gets recorded as 14min to 16min. The difference here is 120s. Hence the max difference is 1min

# Get the difference in seconds, check if withint +-60s
time.taken = journeys$End.Time - journeys$Start.Time
journeys$Time.Valid = time.taken >= (journeys$Journey.Duration - 60) &
  time.taken <= (journeys$Journey.Duration + 60)

# Check that all are valid
head(journeys[!journeys$Time.Valid, ])
```

## Q3

Generally, there is a higher demand for bike sharing on weekdays (Monday to Friday) than on Weekends (Saturday and Sunday). However, it is worth noting that the demand on Wednesday is significantly lower than on other weekdays.

```{r}
# add all the demand of the same day in the same week
journeys$Start.Week = week(journeys$Start.Time)
journeys$Start.Week.Day = weekdays(journeys$Start.Time)
data = journeys %>% group_by(Start.Week, Start.Week.Day) %>%
  summarise(Demand = n())

# Average each day's demand
data = data %>% group_by(Start.Week.Day) %>% summarise(Mean.Demand = mean(Demand))

# Reorder data
order = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
data = data %>% slice(match(order, Start.Week.Day))
data$Start.Week.Day <- factor(data$Start.Week.Day, levels=unique(data$Start.Week.Day))

# plot
ggplot(data=data, aes(x=Start.Week.Day, y=Mean.Demand)) + 
  geom_line(group=1) + 
  geom_point(size=3, shape=23, fill="blue")

```

## Q4

```{r}
# remove brackets
stations$Location = gsub("\\(|\\)", "", stations$Location)
# split cols
stations = stations %>% separate(Location, c("Latitude", "Longitude"), sep=",")

head(stations)
```

## Q5

### (a)

```{r}
# get the start minute of the time slot
journeys$Timeslot =  ifelse(journeys$Start.Minute<30, 0, 30)

# get the start hour
journeys$Timeslot = paste0(journeys$Start.Hour, ":", journeys$Timeslot)

# parse
journeys$Timeslot = strptime(journeys$Timeslot, "%H:%M")

# create end time
journeys$Timeslot = paste0(strftime(journeys$Timeslot, "%H:%M"),
                           "-",
                           strftime(journeys$Timeslot + minutes(30), "%H:%M"))

head(journeys)
```

### (b)

```{r}
journeys$Is.Weekend = journeys$Start.Week.Day %in% c("Saturday", "Sunday")

head(journeys)
```

### (c)

The hypothesis that weekday demand is driven by the practicalities of going to work seem to hold true. The weekday line plot shows higher demands around the 7:30-8:00 timeslot and the 17:30-18:00 timeslot. This means that it is likely people are using the bike sharing to travel to and from work.

The weekend plot, on the other hand, shows a low demand in the early morning and late night and higher demand in the middle of the day, which is what we would expect from leisurely use.

```{r}
# Get the demand for each day
data = journeys %>% group_by(as.Date(Start.Time), Timeslot) %>%
  summarise(Is.Weekend = all(Is.Weekend), Demand = n())

# Get the average demand for weekday/end
data = data %>% group_by(Is.Weekend, Timeslot) %>%
  summarise(Mean.Demand = mean(Demand))

# Plot
ggplot(data=data, aes(x=Timeslot, y=Mean.Demand, group=Is.Weekend, color=Is.Weekend)) + 
  geom_line() + 
  theme(axis.text.x = element_text(angle=90))
```

## Q6

The current planned capacity does not seem to account for the station's crowd. Otherwise, we would see a generally increasing trend. Hence an improvement can be made in the plans to assign higher capacities where there is a greater demand.

```{r}
# Get daily demand
data = journeys %>% group_by(as.Date(Start.Time), Start.Station.ID) %>%
  summarise(Station.Crowd = n())

# Get average crowd per station
data = data %>% group_by(Start.Station.ID) %>%
  summarise(Mean.Station.Crowd = mean(Station.Crowd))

# Merge to get capacity
data = merge(data, stations[c("Station.ID", "Capacity")], 
             by.x="Start.Station.ID", by.y="Station.ID")

# Plot
ggplot(data=data, aes(x=Mean.Station.Crowd, y=Capacity)) + 
  geom_line(group=1)
```

## Q7

```{r}
# Filter for weekdays
data = journeys[!journeys$Is.Weekend,]

# Get the busiest starting stations
busiest.start = journeys %>% group_by(as.Date(Start.Time), Start.Station.ID) %>%
  summarise(Demand = n())
busiest.start = busiest.start %>% group_by(Start.Station.ID) %>%
  summarise(Mean.Demand = mean(Demand))
busiest.start = busiest.start[order(busiest.start$Mean.Demand, decreasing=T), ]
busiest.start = busiest.start[1:10, ]
busiest.start$Busiest.Start = TRUE

# Get the busiest end
busiest.end = journeys %>% group_by(as.Date(End.Time), End.Station.ID) %>%
  summarise(Demand = n())
busiest.end = busiest.end %>% group_by(End.Station.ID) %>%
  summarise(Mean.Demand = mean(Demand))
busiest.end = busiest.end[order(busiest.end$Mean.Demand, decreasing=T), ]
busiest.end = busiest.end[1:10, ]
busiest.end$Busiest.End = TRUE

# join
busiest = merge(busiest.start, busiest.end, 
                by.x=c("Start.Station.ID", "Mean.Demand"), 
                by.y=c("End.Station.ID", "Mean.Demand"), all=T)

head(busiest)
```