---
title: "Assignment 4"
author: "SA01 Team 2"
date: "9/12/2021"
output: html_document
---

```{r,warning=FALSE,message=FALSE}
library(curl)
library(rvest)
library(tidyverse)
library(XML)
```

# Q1

Write a programme to crawl as much information as possible about NBA teams and NBA players from ESPN NBA website: [http://www.espn.com/nba/players](http://www.espn.com/nba/players) .

```{r,warning=FALSE}
url = "http://www.espn.com/nba/players"
page = read_html(url)
nodes = html_nodes(page, ".small-logos div a")

# gets all the url of the different teams
rosters = html_attr(nodes,"href") 
partial = "https://espn.com"
urls = paste0(partial,rosters)

# gets team names
teams = html_text(nodes)

# no. of teams
nteams = length(teams)

# loops through urls and crawls
df = data.frame("Team" = teams)

for (i in 1:nteams)
{
  url = curl(urls[[i]])
  urldata = readLines(url)
  data = readHTMLTable(urldata)
  data$Team = teams[[i]]
  df = merge(df, data, on="Team", all=T)
  close(url)
}

head(df)
```

Cleaning df:

```{r,warning=FALSE}
# handle NA
nba_data = df[!is.na(df$NULL.Name), ]
nba_data[nba_data == "--"] = NA

# column names
nba_data = nba_data[-2]
header = colnames(nba_data) %>% 
  lapply(gsub, pattern="NULL.", replacement="")
names(nba_data) = header

# team number 
nba_data = nba_data %>% separate(Name, 
           into = c("Name", "Jersey"), 
           sep = "(?<=[A-Za-z])(?=[0-9])",
           )

# data type
nba_data$Team = factor(nba_data$Team)
nba_data$POS = factor(nba_data$POS)
nba_data$Jersey = as.integer(nba_data$Jersey)
nba_data$Age = as.integer(nba_data$Age)

nba_data$WT = nba_data$WT %>% 
  gsub(" lbs", "", .) %>%
  as.integer()
names(nba_data)[names(nba_data) == 'WT'] <- 'WT(lbs)'

nba_data$Salary = nba_data$Salary %>% 
  gsub("[$,]", "", .) %>%
  as.numeric()
names(nba_data)[names(nba_data) == 'Salary'] <- 'Salary($)'

# Answer
head(nba_data)
```

# Q2

Based on the discussion today, write a procedure to clean family data (only the gene sequence column). It should include
- A function to validate if a particular input complies with the format requirement;
- A sequence of functions (handlers) to handle all error patterns you could identify in the column.
Run the procedure written above on the given data and make sure you could obtain clean data at the end.

```{r}
family= read_csv("./Data/The family with the largest number of children.csv")
head(family)
```

Check compliance:

```{r}
complies = function(vec) {
  return(grepl("^([FM],)+[FM]$", vec))
}
```

Using validation function to process the data:

```{r}
valid = complies(family$`Gender Sequence`)
family_unclean = family[!valid, ]
family_unclean$original = family_unclean$`Gender Sequence`
family = family[valid, ]
```

Sequence of handlers to handle the error patterns  

1. Convert all to uppercase

```{r}
family_unclean$`Gender Sequence` = toupper(family_unclean$`Gender Sequence`)

# Update validation
valid = complies(family_unclean$`Gender Sequence`)
family = rbind(family, subset(family_unclean[valid, ], select=-c(original)))
family_unclean = family_unclean[!valid, ]
```

2. Convert female/male into m or f

```{r}
family_unclean$`Gender Sequence` = 
  gsub("FEMALE", "F", family_unclean$`Gender Sequence`)
# FMALE vague -> F, Male or Female, handle manually
family_unclean$`Gender Sequence` = 
  gsub("(?<!F)MALE", "M", family_unclean$`Gender Sequence`, perl=T) 

# Update validation
valid = complies(family_unclean$`Gender Sequence`)
family = rbind(family, subset(family_unclean[valid, ], select=-c(original)))
family_unclean = family_unclean[!valid, ]
```

3. Removal of whitespaces

```{r}
family_unclean$`Gender Sequence` = 
  gsub(" ", "", family_unclean$`Gender Sequence`, fixed=T)

# Update validation
valid = complies(family_unclean$`Gender Sequence`)
family = rbind(family, subset(family_unclean[valid, ], select=-c(original)))
family_unclean = family_unclean[!valid, ]
```

4. Drop blank rows

```{r}
family_unclean = family_unclean[!is.na(family_unclean$`Gender Sequence`), ]
```

5. Remove additional symbols

```{r}
family_unclean$`Gender Sequence` = 
  gsub("[^A-Z, ]", "", family_unclean$`Gender Sequence`)

# Update validation
valid = complies(family_unclean$`Gender Sequence`)
family = rbind(family, subset(family_unclean[valid, ], select=-c(original)))
family_unclean = family_unclean[!valid, ]
```

6. Remove additional commas

```{r}
# remove consecutive commas
family_unclean$`Gender Sequence` =
  gsub("[,]+", ",", family_unclean$`Gender Sequence`)
# remove from the ends
family_unclean$`Gender Sequence` = family_unclean$`Gender Sequence` %>%
  gsub("^,", "", .) %>% gsub(",$", "", .)

# Update validation
valid = complies(family_unclean$`Gender Sequence`)
family = rbind(family, subset(family_unclean[valid, ], select=-c(original)))
family_unclean = family_unclean[!valid, ]
```

7. Adding in the commas

```{r}
# find FM and breaks it with ,
select = grepl("^[FM,]+$", family_unclean$`Gender Sequence`) 
family_unclean$`Gender Sequence`[select] = 
  family_unclean$`Gender Sequence`[select] %>%
  gsub("(?<=[FM])(?=[FM])", ",", ., perl=T)

# Update validation
valid = complies(family_unclean$`Gender Sequence`)
family = rbind(family, subset(family_unclean[valid, ], select=-c(original)))
family_unclean = family_unclean[!valid, ]
```

8. Remove Single Child

```{r}
only_child = grepl("^[FM]$", family_unclean$`Gender Sequence`)
family_unclean = family_unclean[!only_child, ]
```

9. Wrong words (Manual intervention)

```{r}
family_unclean # is an only child in this case, no need to add
```

Answer:

```{r}
head(family)
```